Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-02-15 manuscripts/TCS2022/tcs2022.tex
First pass (CP/ILP)

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

@@

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

@@

\linenumbers

@@

% Mention the extensions for the journal version
Herein we revise and extend our previous work in~\cite{DBLP:conf/cmsb/TrinhBHS22} as follows.
First, more formal definitions are given and the existing proofs are made more detailed.
In particular, an updated proof provides another way to prove the independence of trap spaces of a Boolean network on its update scheme, which was originally proved in~\cite{klarner2015computing}.
Second, we showcase a theoretical application of the connection between trap spaces in Boolean networks and conflict-free siphons in Petri nets.
Third, beyond the proposed ASP method implementing the alternative approach~\cite{DBLP:conf/cmsb/TrinhBHS22}, we propose several other possible methods for computing minimal trap spaces using Maximum Satisfiability (MaxSAT), Constraint Logic Programming (CLP), and Integer Linear Programming (ILP).
Fourth, we discuss in detail how to compute several special types of trap spaces in a Boolean network.
Fifth, we present the idea for using our Petri net approach to handle the problem of inconsistent and incomplete data in modeling biological systems.
Sixth, regarding the implementation, we have developed a new converter that directly reads a \texttt{.bnet} file and builds the Petri net encoding, instead of using the \texttt{PNML} conversion of \texttt{bioLQM}~\cite{DBLP:conf/cmsb/TrinhBHS22}.
Finally, we conduct a more comprehensive benchmark on more real-world models from various sources and randomly generated models to evaluate all the proposed methods (the benchmark conducted in~\cite{DBLP:conf/cmsb/TrinhBHS22} considers only real-world models).

@@

\section{Preliminaries}
\label{sec:Preliminaries}

@@

@@

\label{def:BN}
\end{definition}

@@

Note that transforming a Boolean network defined by its functions into its Petri net encoding roughly relies on obtaining conditions for the activation and inactivation of the states. In~\cite{chaouiya2004qualitative} this took the form of the whole truth table of the Boolean functions, but as shown in Appendix 1 of~\cite{chatain2014characterization} computing Disjunctive Normal Forms (DNF) of each Boolean function is enough.
Though this might appear quite computationally intensive it is important to remark first that contrary to the prime-implicants case, there is no need to find \emph{minimal} DNFs.
One way to look at this is to consider that this amounts to a similar approach as that used in~\cite{DBLP:conf/ictai/ChevalierFPZ19} but with the encoding of both activation and inhibition functions as DNFs in order to take into account possible non-local-monotonicity.
This does not change the worst-case-complexity (obtaining a single DNF being exponential) but might matter a lot in practice.
As such, we will explore how this transformation, here using BDDs in \texttt{bioLQM}, and the one based on the most-permissive semantics compare in the Section~\ref{sec:eval} on evaluation.

@@

\begin{proposition}
\label{prop:siphon_set}
  Let \(S\) be a siphon of a Petri net \((P, T, W)\).
  Then \(pred(S) \subseteq succ(S)\).

@@

\section{Minimal trap spaces as maximal conflict-free siphons}
\label{sec:Main_finding}

@@

\section{Computation methods}
\label{sec:Computation}

\subsection{Characterization}
\label{subsec:siphon_characterization}

@@

\subsection{Constraint satisfaction problem}
\label{subsec:computation_csp}

@@

\begin{proposition}
\label{prop:csp_conflict_free_siphon}
  \(\mathcal{C}(\mathcal{P})\) is satisfied by a valuation \(r\) if and only if
  \[
    \{p \in P \;|\; r(p) = 1\}\
  \]
  is a conflict-free siphon of \(\mathcal{P}\).

@@

In~\cite{nabli2016enumerating}, the set of all siphons of a given Petri net is characterized by a similar Boolean CSP except the conflict-freeness constraint.
From the encoded CSP, the set of all \emph{minimal} siphons of the Petri net can be enumerated in the set inclusion order.
For enumerating siphons in the set inclusion order, the proposed method by~\cite{nabli2016enumerating} uses the technique that labels directly the Boolean variables with increasing value selection (i.e., to test first the absence, then the presence of a place in the candidate solution).
The method has two implementations, one uses an iterated SAT procedure and the other uses Constraint Logic Programming (CLP) with backtracking.

One natural question is that how to use the CSP-based method for enumerating all the maximal conflict-free siphons of a Petri net encoding a Boolean network?
Of course, the set of all conflict-free siphons of the Petri net can easily characterized by the CSP model presented in~\cite{nabli2016enumerating} along with the additional constraint \(\neg p_v \vee \neg \overline{p}_v = 1\), for each \(v \in V\), which represents the conflict-freeness.
However, the main concern is to enumerate all the \emph{maximal} ones, which is not trivial to adapt from the CSP-based method.
By Proposition~\ref{prop:csp_conflict_free_siphon}, the set of all maximal conflict-free siphons of \(\mathcal{P}\) can be enumerated in the (maximality) set inclusion order, by restarting the search each time a conflict-free siphon \(S\) is found, with the following additional constraint for disallowing any subset of that conflict-free siphon: \(\bigvee_{p \not \in S} p = 1\).
For enumerating conflict-free siphons in the set inclusion order, we can use the same technique as used in~\cite{nabli2016enumerating} but with the opposite setting, i.e., labeling directly the Boolean variables with decreasing value selection.
The correctness of this technique comes from the fact that once \(S\) is found, it is the conflict-free siphon of maximum cardinality among all the remaining feasible conflict-free siphons.
Similar to~\cite{nabli2016enumerating}, the newly CSP-based method can also be implemented with SAT and CLP solvers.

@@

For the CLP-based method, we use TODO: \dots

\subsection{Answer set programming-based method}
\label{subsec:computation_asp}

@@

\subsection{Integer linear programming-based method}
\label{subsec:computation_ilp}

@@

The conflict-free siphon of maximum cardinality is of course maximal.
Therefore, we impose the following objective function:
\[
  \verb|max| \sum_{v \in V}(\verb|p-v| + \verb|n-v|).
\]
Now, \(\mathcal{I}\) can be solved using a general purpose ILP solver.
If it admits any solution \(I^{*}\), the corresponding conflict-free siphon (say \(S^{*}\)) is maximal.
Hence, it makes sense that it does not need to find any other conflict-free siphon of the net that is strictly contained in \(S^{*}\).
To do this, we add to \(\mathcal{I}\) a new inequality
\[
  \verb|1 <=| \sum_{p \in P \setminus S^{*}} \verb|a-p|
\]
where \verb|a-p| is the binary variable presenting place \(p\).
Now, we solve \(\mathcal{I}\) again to find a new solution.
If a new solution \(I'\) exists, then let \(S'\) be its corresponding conflict-free siphon.
Indeed, abide by the newly added inequality, we have \(S' \cap (P \setminus S^{*}) \neq \emptyset\) because there is some \verb|a-p| with \(p \in P \setminus S^{*}\) such that \(I'(\verb|a-p|) = 1\).
This implies that it is impossible that \(S' = S^{*}\) or \(S' \subset S^{*}\).
By the objective function, it means that \(S'\) is the conflict-free siphon of maximum cardinality among the conflict-free siphons that are not contained in \(S^{*}\).
Hence, \(S'\) is also a maximal conflict-free siphon.
Again, we add to \(\mathcal{I}\) a new inequality with respect to the newly found siphon.
The above process is iterated until \(\mathcal{I}\) becomes unfeasible, this means that there is no further maximal conflict-free siphon.
Thus, all the maximal conflict-free siphons of the Petri net have been found.
We use \dots as the ILP solver. TODO: \dots

\subsection{Computation of special types of trap spaces}
\label{subsec:computation_special_ts}

In the field of systems biology, biologists may want to compute more special types of trap spaces beyond minimal trap spaces~\cite{klarner2017pyboolnet}.
We shall show that our proposed methods can be easily adjusted to compute popular types of trap spaces.
We illustrate the adjustments via the ASP-based method (see Subsection~\ref{subsec:computation_asp}), but these adjustments are completely applicable for other approaches such as MaxSAT, CLP, and ILP.

@@

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v ; n-v.|
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP.

@@

\section{Inconsistent and incomplete data}
\label{sec:incomplete-data}

@@

\label{example:incomplete-BN}
\end{example}

@@

\section{Motivating example}
\label{sec:case_study}

@@

To solve the ASP problems, we used the same ASP solver \texttt{Clingo}~\cite{DBLP:journals/aicom/GebserKKOSS11} and the same configuration as that used in \texttt{PyBoolNet}~\cite{klarner2015computing,klarner2017pyboolnet} and \texttt{mpbn}~\cite{Paulev2020}.
Specifically, we used the configuration \texttt{-heuristic=Domain -enum-mod=domRec -dom-mod=3} (subset maximality, equivalent to the deprecated \texttt{--dom-pref=32 --heuristic=domain --dom-mod=7} used by \texttt{PyBoolNet}). 
We ran all the benchmarks on a machine whose environment is CPU: \@ Intel\textregistered\ Core\texttrademark\ i9-11950H 2.60GHz × 16, 16 GB DDR4 RAM, Ubuntu 20.04.5 LTS. 
Finally, we set a time limit of three minutes for each model.

@@

\setlength{\tabcolsep}{4pt}
\begin{table}[!htb]
  \caption{Timing comparisons (in seconds) between \texttt{bioLQM} (\texttt{LQM}), \texttt{PyBoolNet} (\texttt{PBN}), \texttt{mpbn} and the four variants of \texttt{Trappist} on the \texttt{PyBoolNet} repository.}
  \centering%
  \label{tab:pyboolnet_repo}
  \scalebox{0.85}{
  \begin{tabular}{rlrrrrrrrrr}
    \toprule
    & & & & & & & \multicolumn{4}{c}{\texttt{Trappist}}\\
    \cmidrule(rr){8-11}
    & model & \(n\) & \(|M|\) & LQM & PBN & \texttt{mpbn} & SAT & CLP & ILP & ASP \\
    \midrule
    \rownb & arellano\_rootstem & 9 & 4 & \textbf{0.13} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & calzone\_cellfate & 28 & 27 & \textbf{0.12} & 0.02 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & dahlhaus\_neuroplastoma & 23 & 32 & \textbf{0.11} & 0.03 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & davidich\_yeast & 10 & 12 & \textbf{0.11} & 0.02 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & dinwoodie\_life & 15 & 7 & \textbf{0.11} & 0.01 & 0.00 & 0.01 & - & - & 0.01\\
    \rownb & dinwoodie\_stomatal & 13 & 1 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & faure\_cellcycle & 10 & 2 & \textbf{0.11} & 0.02 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & grieco\_mapk & 53 & 18 & \textbf{0.19} & 0.03 & 0.02 & 0.03 & - & - & 0.02\\
    \rownb & irons\_yeast & 18 & 1 & \textbf{0.12} & 0.03 & 0.01 & 0.01 & - & - & 0.02\\
    \rownb & jaoude\_thdiff & 103 & $>1000$ & N/A & \textbf{0.85} & \textbf{0.45} & \textbf{0.56} & - & - & 0.09\\
    \rownb & klamt\_tcr & 40 & 8 & \textbf{0.11} & 0.01 & 0.01 & 0.01 & - & - & 0.02\\
    \rownb & krumsiek\_myeloid & 11 & 6 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & multivalued & 13 & 4 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n12c5 & 11 & 5 & \textbf{0.11} & \textbf{17.83} & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & n3s1c1a & 2 & 2 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n3s1c1b & 2 & 2 & \textbf{0.09} & 0.02 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n5s3 & 4 & 3 & \textbf{0.10} & 0.02 & \textbf{NM} & 0.00 & - & - & 0.01\\
    \rownb & n6s1c2 & 5 & 3 & \textbf{0.10} & 0.02 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n7s3 & 6 & 3 & \textbf{0.11} & 0.02 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & raf & 3 & 2 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & randomnet\_n15k3 & 15 & 3 & \textbf{0.10} & 0.02 & \textbf{NM} & 0.01 & - & - & 0.01\\
    \rownb & randomnet\_n7k3 & 7 & 10 & \textbf{0.10} & 0.01 & \textbf{NM} & 0.00 & - & - & 0.01\\
    \rownb & remy\_tumorigenesis & 34 & 25 & \textbf{0.15} & \textbf{0.94} & 0.02 & 0.02 & - & - & 0.02\\
    \rownb & saadatpour\_guardcell & 13 & 1 & \textbf{0.10} & 0.06 & 0.00 & 0.00 & - & - & 0.02\\
    \rownb & selvaggio\_emt & 56 & $>1000$ & N/A & \textbf{0.48} & \textbf{0.28} & \textbf{0.28} & - & - & 0.09\\
    \rownb & tournier\_apoptosis & 12 & 3 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & xiao\_wnt5a & 7 & 4 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & zhang\_tlgl & 60 & 156 & \textbf{0.60} & 0.09 & 0.09 & 0.07 & - & - & 0.04\\
    \rownb & zhang\_tlgl\_v2 & 60 & 258 & \textbf{0.64} & 0.04 & 0.08 & 0.11 & - & - & 0.04\\
    \bottomrule
  \end{tabular}
  }
\end{table}
\setcounter{rownumber}{0}

Table~\ref{tab:pyboolnet_repo} shows the experimental results on the models from the official \texttt{PyBoolNet} repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}.
Column \(n\) denotes the number of nodes of each model. 
Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds, asking only for the first 1000 trap spaces.
In the case of \texttt{bioLQM}, ``N/A'' means that the number of all minimal trap spaces of the model is larger than 1000 and we did not recorded the running time of \texttt{bioLQM} because it always requires to compute all minimal trap spaces.
A number in bold indicates a ratio greater than three compared to the best result. 
``NM'' indicates a non-locally-monotonic model.
There are four variants of \texttt{Trappist}: SAT (i.e., the MaxSAT-based method shown in Subsection~\ref{subsec:computation_csp}), CLP (i.e., the CLP-based method shown in Subsection~\ref{subsec:computation_csp}), ILP (i.e., the ILP-based method shown in Subsection~\ref{subsec:computation_ilp}), and ASP (i.e., the ASP-based method shown in Subsection~\ref{subsec:computation_asp}).

@@

\section{Conclusion}
\label{sec:Conclusion}

>>>>

@@

@@

\linenumbers%

@@

% Mention the extensions for the journal version
Herein we revise and extend our previous work in~\cite{DBLP:conf/cmsb/TrinhBHS22} as follows.
First, more formal definitions are given and the existing proofs are made more detailed.
In particular, an updated proof provides another way to prove the independence of trap spaces of a Boolean network on its update scheme, which was originally proved in~\cite{klarner2015computing}.
Second, we showcase a theoretical application of the connection between trap spaces in Boolean networks and conflict-free siphons in Petri nets.
Third, beyond the proposed ASP method implementing the alternative approach~\cite{DBLP:conf/cmsb/TrinhBHS22}, we propose several other possible methods for computing minimal trap spaces using Maximum Satisfiability (MaxSAT), Constraint Programming (CP), and Integer Linear Programming (ILP).
Fourth, we discuss in detail how to compute several special types of trap spaces in a Boolean network.
Fifth, we present the idea for using our Petri net approach to handle the problem of inconsistent and incomplete data in modeling biological systems.
Sixth, regarding the implementation, we have developed a new converter that directly reads a \texttt{.bnet} file and builds the Petri net encoding, instead of using the \texttt{PNML} conversion of \texttt{bioLQM}~\cite{DBLP:conf/cmsb/TrinhBHS22}.
Finally, we conduct a more comprehensive benchmark on more real-world models from various sources and randomly generated models to evaluate all the proposed methods (the benchmark conducted in~\cite{DBLP:conf/cmsb/TrinhBHS22} considers only real-world models).

@@

\section{Preliminaries}%
\label{sec:Preliminaries}

@@

\begin{definition}%
\label{def:BN}

@@

@@

Note that transforming a Boolean network defined by its functions into its Petri net encoding roughly relies on obtaining conditions for the activation and inactivation of the states. In~\cite{chaouiya2004qualitative} this took the form of the whole truth table of the Boolean functions, but as shown in Appendix 1 of~\cite{chatain2014characterization} computing Disjunctive Normal Forms (DNF) of each Boolean function is enough.
Though this might appear quite computationally intensive it is important to remark first that contrary to the prime-implicants case, there is no need to find \emph{minimal} DNFs.
One way to look at this is to consider that this amounts to a similar approach as that used in~\cite{DBLP:conf/ictai/ChevalierFPZ19} but with the encoding of both activation and inhibition functions as DNFs in order to take into account possible non-local-monotonicity.
This does not change the worst-case-complexity (obtaining a single DNF being exponential) but might matter a lot in practice.
As such, we will explore how this transformation, here using BDDs in \texttt{bioLQM} and directly in our tool using the \texttt{pyeda}\footnote{\url{https://pyeda.readthedocs.io/en/latest/}} library, and the one based on the most-permissive semantics compare in the Section~\ref{sec:eval} on evaluation.

@@

\begin{proposition}%
\label{prop:siphon_set}
  Let \(S\) be a siphon of a Petri net \((P, T, W)\).
  Then \(pred(S) \subseteq succ(S)\).

@@

\section{Minimal trap spaces as maximal conflict-free siphons}%
\label{sec:Main_finding}

@@

\section{Computation methods}%
\label{sec:Computation}

\subsection{Characterization}%
\label{subsec:siphon_characterization}

@@

\subsection{Constraint satisfaction problem}%
\label{subsec:computation_csp}

@@

\begin{proposition}%
\label{prop:csp_conflict_free_siphon}
  \(\mathcal{C}(\mathcal{P})\) is satisfied by a valuation \(r\) if and only if
  \[
    \{p \in P \;|\; r(p) = 1\}\
  \]
  is a conflict-free siphon of \(\mathcal{P}\).

@@

In~\cite{nabli2016enumerating}, the set of all siphons of a given Petri net is characterized by a similar Boolean CSP except the conflict-freeness constraint.
From the encoded CSP, the set of all \emph{minimal} siphons of the Petri net can be enumerated in the set inclusion order.
For enumerating siphons in the set inclusion order, the proposed method by~\cite{nabli2016enumerating} uses the technique that labels directly the Boolean variables with increasing value selection (i.e., to test first the absence, then the presence of a place in the candidate solution).
The method has two implementations, one uses an iterated SAT procedure and the other uses Constraint Programming (CP) with backtracking.

One natural question is that how to use the CSP-based method for enumerating all the maximal conflict-free siphons of a Petri net encoding a Boolean network?
Of course, the set of all conflict-free siphons of the Petri net can easily characterized by the CSP model presented in~\cite{nabli2016enumerating} along with the additional constraint \(\neg p_v \vee \neg \overline{p}_v = 1\), for each \(v \in V\), which represents the conflict-freeness.
However, the main concern is to enumerate all the \emph{maximal} ones, which is not trivial to adapt from the CSP-based method.
By Proposition~\ref{prop:csp_conflict_free_siphon}, the set of all maximal conflict-free siphons of \(\mathcal{P}\) can be enumerated in the (maximality) set inclusion order, by restarting the search each time a conflict-free siphon \(S\) is found, with the following additional constraint for disallowing any subset of that conflict-free siphon: \(\bigvee_{p \not \in S} p = 1\).
For enumerating conflict-free siphons in the set inclusion order, we can use the same technique as used in~\cite{nabli2016enumerating} but with the opposite setting, i.e., labeling directly the Boolean variables with decreasing value selection.
The correctness of this technique comes from the fact that once \(S\) is found, it is the conflict-free siphon of maximum cardinality among all the remaining feasible conflict-free siphons.
Similar to~\cite{nabli2016enumerating}, the newly CSP-based method can also be implemented with SAT and CP solvers.

This method was implemented using the state-of-the-art CP solver Chuffed\footnote{\url{https://github.com/chuffed/chuffed}}~\cite{demirovic2018solution} via its MiniZinc~\cite{nethercote2007minizinc} interface.
Because it is a high-level interface, the backtrack-and-replay method of~\cite{nabli2016enumerating} was not used but rather the alternative implementation with two global constraints for lexicographic ordering (ensuring enumeration of solutions) and iterated non-subset of each already found solution (for maximality).

@@

\subsection{Answer set programming-based method}%
\label{subsec:computation_asp}

@@

\subsection{Integer linear programming-based method}%
\label{subsec:computation_ilp}

@@

The conflict-free siphon of maximum cardinality is of course maximal.
Therefore, we impose the following objective function:
\[
  \verb|max| \sum_{v \in V}(\verb|p-v| + \verb|n-v|).
\]
Now, \(\mathcal{I}\) can be solved using a general purpose ILP solver.
If it admits any solution \(I^{*}\), the corresponding conflict-free siphon (say \(S^{*}\)) is maximal.
Hence, it makes sense that it does not need to find any other conflict-free siphon of the net that is strictly contained in \(S^{*}\).
To do this, we add to \(\mathcal{I}\) a new inequality
\[
  \verb|1 <=| \sum_{p \in P \setminus S^{*}} \verb|a-p|
\]
where \verb|a-p| is the binary variable presenting place \(p\).
Now, we solve \(\mathcal{I}\) again to find a new solution.
If a new solution \(I'\) exists, then let \(S'\) be its corresponding conflict-free siphon.
Indeed, abide by the newly added inequality, we have \(S' \cap (P \setminus S^{*}) \neq \emptyset\) because there is some \verb|a-p| with \(p \in P \setminus S^{*}\) such that \(I'(\verb|a-p|) = 1\).
This implies that it is impossible that \(S' = S^{*}\) or \(S' \subset S^{*}\).
By the objective function, it means that \(S'\) is the conflict-free siphon of maximum cardinality among the conflict-free siphons that are not contained in \(S^{*}\).
Hence, \(S'\) is also a maximal conflict-free siphon.
Again, we add to \(\mathcal{I}\) a new inequality with respect to the newly found siphon.
The above process is iterated until \(\mathcal{I}\) becomes unfeasible, this means that there is no further maximal conflict-free siphon.
Thus, all the maximal conflict-free siphons of the Petri net have been found.

Since we used the MiniZinc framework to interface with the CP solver, it was simple to make the slight modifications described above and use that same interface to call the Coin-OR CBC solver\footnote{\url{https://github.com/coin-or/Cbc}}~\cite{john_forrest_2022_6522795}.

\subsection{Computation of special types of trap spaces}%
\label{subsec:computation_special_ts}

In the field of systems biology, biologists may want to compute more special types of trap spaces beyond minimal trap spaces~\cite{klarner2017pyboolnet}.
We shall show that our proposed methods can be easily adjusted to compute popular types of trap spaces.
We illustrate the adjustments via the ASP-based method (see Subsection~\ref{subsec:computation_asp}), but these adjustments are completely applicable for other approaches such as MaxSAT, CP, and ILP\@.

@@

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v ; n-v.|
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP\@.

@@

\section{Inconsistent and incomplete data}%
\label{sec:incomplete-data}

@@

\label{example:incomplete-BN}%
\end{example}

@@

\section{Motivating example}%
\label{sec:case_study}

@@

To solve the ASP problems, we used the same ASP solver \texttt{Clingo}~\cite{DBLP:journals/aicom/GebserKKOSS11} and the same configuration as that used in \texttt{PyBoolNet}~\cite{klarner2015computing,klarner2017pyboolnet} and \texttt{mpbn}~\cite{Paulev2020}.
Specifically, we used the configuration \texttt{-heuristic=Domain -enum-mod=domRec -dom-mod=3} (subset maximality, equivalent to the deprecated \texttt{--dom-pref=32 --heuristic=domain --dom-mod=7} used by \texttt{PyBoolNet}).
We ran all the benchmarks on a machine whose environment is CPU\@: Intel\textregistered\ Core\texttrademark\ i9-11950H 2.60GHz × 16, 16 GB DDR4 RAM, Ubuntu 20.04.5 LTS\@.
Finally, we set a time limit of three minutes for each model.

@@

\setlength{\tabcolsep}{4pt}
\begin{table}[!htb]
  \caption{Timing comparisons (in seconds) between \texttt{bioLQM} (\texttt{LQM}), \texttt{PyBoolNet} (\texttt{PBN}), \texttt{mpbn} and the four variants of \texttt{Trappist} on the \texttt{PyBoolNet} repository.}
  \centering%
  \label{tab:pyboolnet_repo}
  \scalebox{0.85}{
  \begin{tabular}{rlrrrrrrrrr}
    \toprule
    & & & & & & & \multicolumn{4}{c}{\texttt{Trappist}}\\
    \cmidrule(rr){8-11}
    & model & \(n\) & \(|M|\) & LQM & PBN & \texttt{mpbn} & SAT & CP & ILP & ASP \\
    \midrule
    \rownb & arellano\_rootstem & 9 & 4 & \textbf{0.13} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & calzone\_cellfate & 28 & 27 & \textbf{0.12} & 0.02 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & dahlhaus\_neuroplastoma & 23 & 32 & \textbf{0.11} & 0.03 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & davidich\_yeast & 10 & 12 & \textbf{0.11} & 0.02 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & dinwoodie\_life & 15 & 7 & \textbf{0.11} & 0.01 & 0.00 & 0.01 & - & - & 0.01\\
    \rownb & dinwoodie\_stomatal & 13 & 1 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & faure\_cellcycle & 10 & 2 & \textbf{0.11} & 0.02 & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & grieco\_mapk & 53 & 18 & \textbf{0.19} & 0.03 & 0.02 & 0.03 & - & - & 0.02\\
    \rownb & irons\_yeast & 18 & 1 & \textbf{0.12} & 0.03 & 0.01 & 0.01 & - & - & 0.02\\
    \rownb & jaoude\_thdiff & 103 & $>1000$ & N/A & \textbf{0.85} & \textbf{0.45} & \textbf{0.56} & - & - & 0.09\\
    \rownb & klamt\_tcr & 40 & 8 & \textbf{0.11} & 0.01 & 0.01 & 0.01 & - & - & 0.02\\
    \rownb & krumsiek\_myeloid & 11 & 6 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & multivalued & 13 & 4 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n12c5 & 11 & 5 & \textbf{0.11} & \textbf{17.83} & 0.01 & 0.01 & - & - & 0.01\\
    \rownb & n3s1c1a & 2 & 2 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n3s1c1b & 2 & 2 & \textbf{0.09} & 0.02 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n5s3 & 4 & 3 & \textbf{0.10} & 0.02 & \textbf{NM} & 0.00 & - & - & 0.01\\
    \rownb & n6s1c2 & 5 & 3 & \textbf{0.10} & 0.02 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & n7s3 & 6 & 3 & \textbf{0.11} & 0.02 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & raf & 3 & 2 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & randomnet\_n15k3 & 15 & 3 & \textbf{0.10} & 0.02 & \textbf{NM} & 0.01 & - & - & 0.01\\
    \rownb & randomnet\_n7k3 & 7 & 10 & \textbf{0.10} & 0.01 & \textbf{NM} & 0.00 & - & - & 0.01\\
    \rownb & remy\_tumorigenesis & 34 & 25 & \textbf{0.15} & \textbf{0.94} & 0.02 & 0.02 & - & - & 0.02\\
    \rownb & saadatpour\_guardcell & 13 & 1 & \textbf{0.10} & 0.06 & 0.00 & 0.00 & - & - & 0.02\\
    \rownb & selvaggio\_emt & 56 & $>1000$ & N/A & \textbf{0.48} & \textbf{0.28} & \textbf{0.28} & - & - & 0.09\\
    \rownb & tournier\_apoptosis & 12 & 3 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & xiao\_wnt5a & 7 & 4 & \textbf{0.10} & 0.01 & 0.00 & 0.00 & - & - & 0.01\\
    \rownb & zhang\_tlgl & 60 & 156 & \textbf{0.60} & 0.09 & 0.09 & 0.07 & - & - & 0.04\\
    \rownb & zhang\_tlgl\_v2 & 60 & 258 & \textbf{0.64} & 0.04 & 0.08 & 0.11 & - & - & 0.04\\
    \bottomrule
  \end{tabular}
  }
\end{table}
\setcounter{rownumber}{0}

Table~\ref{tab:pyboolnet_repo} shows the experimental results on the models from the official \texttt{PyBoolNet} repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}.
Column \(n\) denotes the number of nodes of each model.
Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds, asking only for the first 1000 trap spaces.
In the case of \texttt{bioLQM}, ``N/A'' means that the number of all minimal trap spaces of the model is larger than 1000 and we did not recorded the running time of \texttt{bioLQM} because it always requires to compute all minimal trap spaces.
A number in bold indicates a ratio greater than three compared to the best result.
``NM'' indicates a non-locally-monotonic model.
There are four variants of \texttt{Trappist}: SAT (i.e., the MaxSAT-based method shown in Subsection~\ref{subsec:computation_csp}), CP (i.e., the CP-based method shown in Subsection~\ref{subsec:computation_csp}), ILP (i.e., the ILP-based method shown in Subsection~\ref{subsec:computation_ilp}), and ASP (i.e., the ASP-based method shown in Subsection~\ref{subsec:computation_asp}).

@@

\section{Conclusion}%
\label{sec:Conclusion}

In this article we have explored and proved for the first time the equivalence between (minimal) trap spaces of a general Boolean network and (maximal) conflict-free siphons of its Petri net encoding.
We have shown several important applications of this finding to studying properties of trap spaces in Boolean networks.
As an important practical application of the equivalence, we have proposed a new approach for the computation of minimal trap spaces of Boolean networks, based on the enumeration of maximal conflict-free siphons of Petri nets.
We have also proposed the four possible methods using MaxSAT, CP, ILP, and ASP for implementing the new approach.
The proposed methods have been evaluated on many real-world models from the literature and randomly generated models.
The experimental results show that the new approach vastly outperforms all the state-of-the-art methods in terms of general Boolean networks and is comparable to the \texttt{mpbn} method even better in average in term of locally-monotonic Boolean networks.
We believe that this opens up the way to a much better analysis of large Boolean networks, which is needed with the advent of automatic model-generation pipelines~\cite{ostaszewski2021covid19}.
